 

### 2021-06-10 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

However, when it comes to sun and shade leaves, a few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] and other studies have observed no difference in $Q_{10}$ across species vertical profiles [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999], and at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered soil-water dynamics than milder drought, particulary in temperate species [@ruehrWaterAvailabilityDominant2016]. However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate]. Further, heat with drought exacerbates tree stress [@ruehrWaterAvailabilityDominant2016]. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al. 2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].[**to include Teskey 2015; Breashers et al. 2021**]

>>>>

However, when it comes to sun and shade leaves, a few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] and other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], and at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Accordingly, in temperate species, heat stress caused due to increase in temperature and VPD even under adequete water caused greater tree hydraulic stress and altered soil-water dynamics than milder drought [@ruehrWaterAvailabilityDominant2016]. However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate]. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al. 2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018]. Further, warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
[**to include Teskey 2015; Breashers et al. 2021**]
